One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-07-31 papers/1_dmatpol/paper-r3.tex
Figures and Tables revised.

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities.
The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially uniform electric field, as well as
in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The electric field\hyp{}induced polarization energies,
dipole moments and quadrupole moments are quantitatively described by the present theory 
in weak, moderate and even very strong electric fields (0.003 -- 0.07 a.u.)
when quadratic terms with respect to the electric field are accounted for.
It is believed that the proposed model could open
new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
\end{abstract}

@@

However, even with the modern advances in computational chemistry, 
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,
Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017} 
Interface between QM and MM levels is usually realized by electrostatic embedding
and only rarely includes the Pauli exclusion principle and dispersion.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
because of complicated nature of intermolecular interactions, that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Perhaps the most frequently used efficient and reasonably accurate \emph{ab initio} fragmentation methods,
the effective fragment potential (EFP) methods, are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF)\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.

@@

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the follwoing Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element could accumulate tremendously when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are probably not sufficient enough to accurately describe the local polarization
of electron density distribution when electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed\hyp{}site model at the Hartree\hyp{}Fock (HF)
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, 
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C} \cdot {\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right] \cdot
                                   \left[ {\bf 1} + {\bf v}^\dagger \cdot {\bf v} \right]^{-1} \cdot
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)} + 
                        {\bf D}^{(0)} \cdot {\BM\Delta}^\dagger \cdot \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C} \cdot {\bf C}^{(0)\dagger} + {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \nonumber \\
           &\qquad - {\bf D}^{(0)} \cdot \delta {\bf C} \cdot {\bf C}^{{(0)}\dagger} 
                   - {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \cdot {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger} \cdot {\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated otherwise.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This guarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;.
 \end{align}
\end{subequations}
%
to simplify the notation. Hence, we have
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let us partition the induced dipole moment into separate contributions
from the occupied MO's, i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed polarizability tensor
associated with the $i$th MO whereas ${\bf F}({\bf r}_i)$ is the electric field evaluated at the $i$th MO centroid of charge
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
We can now insert the above result into Eq.~\eqref{e:dmu-l-vector} 
and after recasting the right hand side in a matrix form
we get
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of size $(n \times 3)$ and $\delta{\bf c}_i$ is just the $i$th column of the
change in the LCAO-MO matrix due to the polarization process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix 
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T 
\end{equation}
%
and has size $(3\times n)$. Note that the \emph{right inverse} of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarization.
Quadratic terms with respect to the electric field as well as linear terms with respect to
electric field gradient can be added to the formulation, which leads to
%
\begin{multline}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \Big\{
                                      {\bf B}^{(i ; 1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \\
%                       + \sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) 
                        + \ldots \Big\}
\end{multline}
%
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; R)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field.
Note that
${\bf B}^{(i; 1)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 2)}_{\alpha\beta}$
is $(3\times 3)$ matrix. Also, the distributed expansion centres
(labeled as $i$) are generalized to any point in the molecule.\footnote{More general
contribution due to square of electric field could be $\sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)$. However, for simplicity we neglect the contributions from different sites $i$ and $j$
in the present discussion and leave it for future development.
}

@@

\begin{figure*}[h]
\includegraphics[width=\textwidth]{data/dmatpol/water/figure1/fig-2.eps}
\caption{\label{f:fig-2} {\bf Performance of the distributed density matrix susceptibility tensors in moderate
non\hyp{}uniform electric fields.}
Calculations of the electric field\hyp{}induced interaction energies (A-C), dipole moments (D-F) and
quadrupole moments (G-I). Comparison with the distributed polarizability model is also shown in this Figure
in panels C, F and I.}
\end{figure*}

>>>>

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities (DMS)
interacting with electric field.
The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially uniform electric field, as well as
in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The electric field\hyp{}induced polarization energies,
dipole moments and quadrupole moments are quantitatively described by the present theory 
in weak, moderate and even very strong electric fields (0.003 -- 0.07 a.u.)
when quadratic terms with respect to the electric field are accounted for.
It is believed that the proposed model could open
new routes in Quantum Chemistry of fast and efficient calculations of strongly perturbed systems.
\end{abstract}

@@

However, even with the modern advances in computational chemistry, 
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,
Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017} 
Interface between QM and MM levels is usually realized by electrostatic embedding
and only rarely includes the Pauli exclusion principle and dispersion.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
because of complicated nature of intermolecular interactions that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Perhaps one of the most frequently used, efficient and reasonably accurate \emph{ab initio} fragmentation methods,
the effective fragment potential (EFP) methods, are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF)\cite{Roothaan.RevModPhys.1951,Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.

@@

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the follwoing Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element could accumulate tremendously when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are probably not sufficient enough to accurately describe the local polarization
of electron density distribution when electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed\hyp{}site model at the Hartree\hyp{}Fock
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, 
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is atomic orbital (AO)\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C} \cdot {\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals (MO's) as linear combinations
of AO's, referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right] \cdot
                                   \left[ {\bf 1} + {\bf v}^\dagger \cdot {\bf v} \right]^{-1} \cdot
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)} + 
                        {\bf D}^{(0)} \cdot {\BM\Delta}^\dagger \cdot \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C} \cdot {\bf C}^{(0)\dagger} + {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \nonumber \\
           &\qquad - {\bf D}^{(0)} \cdot \delta {\bf C} \cdot {\bf C}^{{(0)}\dagger} 
                   - {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \cdot {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger} \cdot {\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated otherwise.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This guarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;.
 \end{align}
\end{subequations}
%
to simplify the notation. Hence, we have
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let us partition the induced dipole moment into separate contributions
from the occupied MO's, i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarizability tensor
associated with the $i$th MO whereas ${\bf F}({\bf r}_i)$ is the electric field evaluated at the $i$th MO centroid of charge,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
We can now insert the above result into Eq.~\eqref{e:dmu-l-vector} 
and after recasting the right hand side in a matrix form
and assuming real quantities
we get
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of size $(n \times 3)$ and $\delta{\bf c}_i$ is just the $i$th column of the
change in the LCAO-MO matrix due to the polarization process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T   \;,
\end{equation}
%
and has size $(3\times n)$. Note that the \emph{right inverse} of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarization.
Quadratic (and higher\hyp{}order) terms with respect to the electric field
can be added to the formulation, which leads to
%
\begin{multline}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \Big\{
                                      {\bf B}^{(i ; 1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \\
%                       + \sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) 
                        + \ldots \Big\}
\end{multline}
%
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; R)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field.
Note that
${\bf B}^{(i; 1)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 2)}_{\alpha\beta}$
is a $(3\times 3)$ matrix. Also, the distributed expansion centres
(labeled as $i$) are generalized to any point in the molecule.\footnote{More general
contribution due to square of electric field could 
be $\sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)$. 
However, for simplicity, we neglect the contributions from different sites $i$ and $j$
in the present discussion and leave it for future development.
}

@@

\begin{figure*}[h]
\includegraphics[width=\textwidth]{data/dmatpol/water/figure1/fig-2.eps}
\caption{\label{f:fig-2} {\bf Performance of the distributed DMS tensors of water molecule in moderate
non\hyp{}uniform electric fields.}
Calculations of the electric field\hyp{}induced interaction energies (A-C), dipole moments (D-F) and
quadrupole moments (G-I). Comparison with the distributed polarizability model is also shown in this Figure
in panels C, F and I.}
\end{figure*}
